Spatial Reasoning and Statistical Graphics

### 2014-10-16 Paper/Draft1/TestingVisualAptitude.Rnw
Move comment, add more explanation

\begin{savenotes}
\begin{table}[ht]
\centering
\begin{tabular}{rlll}
  & Card Rotation & Paper Folding & Figure Classification \\\hline
ISU Students & \Sexpr{round(mean(ans.summary.unscaled$card_rot), 2)} (\Sexpr{round(sd(ans.summary.unscaled$card_rot), 1)}) 
             & \Sexpr{round(mean(ans.summary.unscaled$folding), 2)} (\Sexpr{round(sd(ans.summary.unscaled$folding), 1)})
             & \Sexpr{round(mean(ans.summary.unscaled$fig_class), 2)} (\Sexpr{round(sd(ans.summary.unscaled$fig_class), 1)})\footnote{ISU students took only Part I due to time constraints.}\\
Scaled Scores & 88.0 (\Sexpr{round(sqrt((24.6^2)*2), 1)})
              & 13.8 (4.5)
              & \Sexpr{round((120.0*294 + 114.9*323)/617/2, 2)} (\Sexpr{round(sqrt((30^2*294 + 27.8^2*323)/617/2), 1)})\footnote{Averages calculated assuming 294 males and 323 females.}\\
Unscaled Scores & 44.0 (24.6)\footnote{Data from Part I only.}
                & 13.8 (4.5)
                & M:~120.0~(30.0), F:~114.9~(27.8)\\
{\footnotesize Population}    
              & \parbox[t]{.2\textwidth}{\footnotesize $\approx$ 550 male Naval recruits} 
              & \parbox[t]{.2\textwidth}{\footnotesize 46 college students\\(1963~version)}
              & \parbox[t]{.3\textwidth}{\footnotesize Suburban 11-12th grade students\\(288-300 males, 317-329 females)}\\\hline
\end{tabular}

@@

\newtext{Table \ref{tab:scorecomparison} shows mean scores and standard deviation for ISU students and other populations compared using the Kit of Factor-Referenced Cognitive Tests \citet{ekstrom1976manual}. Once values have been adjusted to accommodate different test procedures (some data is reported for a single part of a two-part test, for instance), it is evident that Iowa State undergraduates scored at about the same level as other similar demographics. In fact, both means and standard deviations of ISU students' scores are similar to the comparison groups, which were chosen from available demographic groups based on population similarity.}\footnote{That is, if comparison data was available for 9th grade students and 12th grade students, we have compared Iowa State students' scores with the 12th grade students.}

\newtext{For the remainder of this analysis, we will scale the test results so that the ranges and units of test scores are comparable.} Let $X_{n, k}$ be a participant's score on a test consisting of $n$ questions with $k$ answers each, out of which only one is correct.
This leads to a theoretical range of $X_{n,k}$ of $[-n/(k-1), n]$ and, under an additional assumption of random guessing, a variance of

\begin{eqnarray*}
\text{Var}(X_{n, k}) &=& n^2 \text{Var}(X_{1, k}) = \\
&=& n^2 (\underbrace{1/k \cdot 1^2}_{\text{correct answer}}+ \underbrace{(-1/(k-1))^2 \cdot (k-1)/k}_{\text{wrong answer}}) = \\
&=& n^2/(k-1).\\
\text{Sd}(X_{n, k}) &=& n/sqrt{k-1}
\end{eqnarray*}

@@

<<correlations, fig.cap="Scree plot of principle component analysis of performance on the different test batteries.", out.width=".5\\linewidth", fig.height=4.5, fig.width=7.5, comment="|", warning=F, message=F>>=
cor(ans.summary[,c("lineup","card_rot", "fig_class", "folding", "vis_search")])
# using scaled version right now, should be changed to unscaled once the scores are internally scaled.
pca <- prcomp(ans.summary[,c("lineup", "card_rot", "fig_class", "folding", "vis_search")], scale=F)
summary(pca)
screeplot(pca)
pca$rotation

# Just using the 4 cognitive tests and ignoring the lineups...
pca.cog.tests <- prcomp(ans.summary[,c("card_rot", "fig_class", "folding", "vis_search")])
summary(pca.cog.tests)
screeplot(pca.cog.tests)
pca.cog.tests$rotation
@

\comment{If test scores have the same ranges, we could include a mini table of means and standard deviations of the four tests, and discuss whether these findings are consistent to how people usually score (except for the females we have about the same population). \textit{I'm not sure where to find the ``normal scores" in the factor-referenced manual. I'll check through it again, but I don't recall seeing it.}}

>>>>

\newtext{The advantage of using tests from the Kit of Factor Referenced Cognitive tests \citep{ekstrom1976manual} is that the tests are extremely well studied \citep{voyer1995manual} and comparison data are available from the validation of these factors. The card rotation, paper folding, and figure classification tests have been validated using different populations, many of which are demographically similar to Iowa State students (naval recruits, college students, late high-school students, and 9th grade students).}

% \comment{If test scores have the same ranges, we could include a mini table of means and standard deviations of the four tests, and discuss whether these findings are consistent to how people usually score (except for the females we have about the same population).}
\begin{savenotes}
\begin{table}[ht]
\centering
\begin{tabular}{rlll}
  & Card Rotation & Paper Folding & Figure Classification \\\hline
ISU Students & \Sexpr{round(mean(ans.summary.unscaled$card_rot), 2)} (\Sexpr{round(sd(ans.summary.unscaled$card_rot), 1)}) 
             & \Sexpr{round(mean(ans.summary.unscaled$folding), 2)} (\Sexpr{round(sd(ans.summary.unscaled$folding), 1)})
             & \Sexpr{round(mean(ans.summary.unscaled$fig_class), 2)} (\Sexpr{round(sd(ans.summary.unscaled$fig_class), 1)})\footnote{ISU students took only Part I due to time constraints.}\\
Scaled Scores & 88.0 (\Sexpr{round(sqrt((24.6^2)*2), 1)})
              & 13.8 (4.5)
              & \Sexpr{round((120.0*294 + 114.9*323)/617/2, 2)} (\Sexpr{round(sqrt((30^2*294 + 27.8^2*323)/617/2), 1)})\footnote{Averages calculated assuming 294 males and 323 females.}\\
Unscaled Scores & 44.0 (24.6)\footnote{Data from Part I only.}
                & 13.8 (4.5)
                & M:~120.0~(30.0), F:~114.9~(27.8)\\
{\footnotesize Population}    
              & \parbox[t]{.2\textwidth}{\footnotesize $\approx$ 550 male Naval recruits} 
              & \parbox[t]{.2\textwidth}{\footnotesize 46 college students\\(1963~version)}
              & \parbox[t]{.3\textwidth}{\footnotesize Suburban 11-12th grade students\\(288-300 males, 317-329 females)}\\\hline
\end{tabular}

@@

\newtext{Table \ref{tab:scorecomparison} shows mean scores and standard deviation for ISU students and other populations. Once values have been adjusted to accommodate different test procedures (some data is reported for a single part of a two-part test, for instance), it is evident that Iowa State undergraduates scored at about the same level as other similar demographics. In fact, both means and standard deviations of ISU students' scores are similar to the comparison groups, which were chosen from available demographic groups based on population similarity.}\footnote{That is, if comparison data was available for 9th grade students and 12th grade students, we have compared Iowa State students' scores with the 12th grade students, as they are closer in age to college students. When data was available from college students and Army enlistees, we have compared ISU students to other college students, as college students are more likely to have similar gender distribution to ISU students.}

\newtext{As we have established that the results obtained for the ETS tests are similar to other studies, we will now compare the results to the lineups also tested in this study. To facilitate this goal, for the remainder of this analysis, we will scale the test results so that the ranges and units of test scores are comparable.} Let $X_{n, k}$ be a participant's score on a test consisting of $n$ questions with $k$ answers each, out of which only one is correct.
This leads to a theoretical range of $X_{n,k}$ of $[-n/(k-1), n]$ and, under an additional assumption of random guessing, a variance of

\begin{eqnarray*}
\text{Var}(X_{n, k}) &=& n^2 \text{Var}(X_{1, k}) = \\
&=& n^2 (\underbrace{1/k \cdot 1^2}_{\text{correct answer}}+ \underbrace{(-1/(k-1))^2 \cdot (k-1)/k}_{\text{wrong answer}}) = \\
&=& n^2/(k-1).
\end{eqnarray*}

@@

<<correlations, fig.cap="Scree plot of principle component analysis of performance on the different test batteries.", out.width=".5\\linewidth", fig.height=4.5, fig.width=7.5, comment="|", warning=F, message=F>>=
cor(ans.summary[,c("lineup","card_rot", "fig_class", "folding", "vis_search")])
# using scaled version right now, should be changed to unscaled once the scores are internally scaled.
pca <- prcomp(ans.summary[,c("lineup", "card_rot", "fig_class", "folding", "vis_search")], scale=F)
summary(pca)
screeplot(pca)
pca$rotation
@

<<pca, fig.cap="Plots of the principle components with observations. Visual Search and Paper Folding strongly contribute to both PC1 and PC2, while Figure Classification and Paper Folding strongly contributes to PC3 and Card Rotation strongly contributes to PC4", comment="|", fig.height=4.5, fig.width=4.5, warning=F, message=F, out.width='.45\\textwidth'>>=
# Just using the 4 cognitive tests and ignoring the lineups...
pca.cog.tests <- prcomp(ans.summary[,c("card_rot", "fig_class", "folding", "vis_search")], retx=T)
summary(pca.cog.tests)
pca.cog.tests$rotation
biplot(pca.cog.tests, choices=1:2, pc.biplot=T, cex=c(.5, 1), adj=.75)
biplot(pca.cog.tests, choices=3:4, pc.biplot=T, cex=c(.5, 1), adj=.75)
@

<<pca-reg, comment="|", warning=F, message=F>>=
# Linear regression using PC's. 
ans.pca <- cbind(ans.summary[,c(1:19)], ans.summary["lineup"], pca.cog.tests$x)
summary(lm(lineup~PC1+PC2+PC3+PC4, data=ans.pca))
@
